---
title: "431 Lab 04"
date: "Last Edited `r Sys.time()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# Instructions

Lab 04 includes 8 questions. Be sure to respond to each question by the deadline posted on the [Course Calendar](https://thomaselove.github.io/431/calendar.html). 

You are welcome (encouraged, really) to discuss Lab 04 with Dr. Love, the teaching assistants and even your colleagues, but your answer must be prepared by you alone. Don't be afraid to ask questions, either via [Piazza](https://piazza.com/case/fall2020/pqhs431) (use the lab04 folder), at [TA office hours](https://thomaselove.github.io/431/contact.html) or before/after class.

Your response to **every** question, whether we explicitly ask for it or not, should include a complete English sentence responding to the question. Code alone is not a sufficient response, even if the code is correct. Some questions, of course, don't require any code.

We have not provided a template for this Lab. You are encouraged to adapt one of the previous templates for this purpose.

## Submitting your Response

Build your entire response as an R Markdown file. Then knit the file to create an HTML document. Submit both your R Markdown file and the HTML output file to [Canvas in the Lab 04 section of the Assignments folder](https://canvas.case.edu) by the deadline specified in [the Course Calendar](https://thomaselove.github.io/431/calendar.html).

## Minimal Setup

The minimal setup for this assignment involves loading two relevant R packages, as shown below. In our answer sketch we may add additional packages, and you are welcome to do so in your response, as well. Remember to load the tidyverse last.

```{r, message = FALSE}
library(palmerpenguins); library(tidyverse)
## make sure these libraries are installed in R
## always need tidyverse, also need palmerpenguins for Lab 04
```

# Setup for Questions 1-5

Find a headline from the internet related to health or medicine that describes the findings of a study published on January 1, 2017 or later. Then find the study being referred to in PUBMED. Use the [formula for updating your opinions about health news developed by Jeff Leek](http://fivethirtyeight.com/features/a-formula-for-decoding-health-news/), along with the abstract and full contents of the published study to complete Questions 1-5. While it won't be necessary to prepare any R code to respond to Questions 1-5, we think it will be good practice for you to prepare your response in R Markdown anyway.

# Question 1 (5 points)

Specify the URL where we can see the headline and news story describing the findings of the study. Feel free to use `bit.ly` or a related tool online to produce a shortened URL for this purpose. Specify the reference completely, including the names of the author(s) of the news story, and its full title, and source.

# Question 2 (5 points)

Specify a URL where we can see at least the abstract of the complete study. Again, shortened URLs are fine. Give the complete reference to the study, as well, including the authors, full title, journal name and so forth. 

# Question 3 (10 points)

Describe, in a few sentences, your original opinion (gut feeling) related to the conclusions of the study as summarized in the headline and news article, first in terms of a probability statement, and then calculate the appropriate odds, remembering to convert statements about probabilities to statements about odds. Provide some motivation for your internal prior probability, describing your relevant personal experiences or other factors that drove your gut feeling.

Remember, if X is an event, and Pr(X) is the probability that X occurs, and odds(X) are the odds that X occurs, then 

**Pr(X) = odds(X) / [1 + odds(X)]**

and 

**odds(X) = Pr(X) / (1 - Pr(X))**.

# Question 4 (20 points)

Evaluate the study in terms of the six specifications [proposed by Leek](http://fivethirtyeight.com/features/a-formula-for-decoding-health-news/) when evaluating study support. Be sure to specify your conclusion about **each** of the six specifications, and provide direct quotes and summarize the evidence from the abstract or paper to address the issues raised and justify your conclusions.

# Question 5 (20 points)

Incorporate the study support assessment into a Bayes' Rule calculation to obtain the final odds you should now be willing to give to the headline, and specify this value in terms of a probability statement, as well. Then react to the final conclusion specified by this approach in a few sentences. How does your subjective posterior probability that the headline is true match up with the formula's conclusions? Do you feel that the formulaic approach has yielded an appropriate conclusion for you in this case? Why or why not?

# Setup for Questions 6-8

We'll be using the `penguins` data (note: use the `penguins` tibble, and not the `penguins_raw` tibble for this Lab) contained in the `palmerpenguins` package in R. The complete citation is ...

Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago (Antarctica) penguin data. R package version 0.1.0. https://allisonhorst.github.io/palmerpenguins/. doi: 10.5281/zenodo.3960218.

Additional information on the data are provided:

- by Allison Horst at the github site cited above. In particular, you'll find a nice cartoon of [the three species of penguin contained in the data](https://github.com/allisonhorst/palmerpenguins#meet-the-palmer-penguins) and a detailed [description of the bill measurements](https://github.com/allisonhorst/palmerpenguins#bill-dimensions) that are worth your time, and
- in [Chapter 2 of the Course Notes](https://thomaselove.github.io/431-notes/looking-at-the-palmer-penguins.html).

You'll be building plots in Questions 6-8, and we want you to use the tidyverse (specifically the ggplot2 package) to do so. 

# Question 6 (15 points)

Suppose we are interested in which of the three species of penguin (*Adelie*, *Chinstrap* or *Gentoo*) shows the strongest *linear* relationship between bill length and body mass, now using all 342 penguins with complete data on the two variables we're studying (bill length and body mass.) 

Draw an attractive and thoughtfully labeled plot (including a title) to show the association of bill length (placed on the horizontal x axis) and body mass (on the vertical y axis) for these 342 penguins, and facet the plot by species (so that one facet shows the 151 Adelie, one shows the 68 Chinstrap and one shows the 123 Gentoo.) Use `geom_smooth()` to add an appropriate smooth curve (using `lm` or `loess` or both, as you prefer) to each facet of the plot.

Then write a sentence or two describing what your plot reveals.

# Question 7 (15 points)

Build a linear model predicting body mass using bill length for the species which has the largest body mass (on average) across the three species. 

Then use that equation to estimate the difference in the predicted body mass for two new penguins of that species (named Pingu and Skipper), where Pingu whom has a bill length at the 75th percentile of the original data for that species, and Skipper has a bill length at the 25th percentile of the original data for that species? Maintain two decimal places for any intermediate calculations, although your final response should be provided in grams (as an integer.)

In a summary sentence or two describing your results, be sure to specify which species you are studying and which of the two new penguins (Pingu or Skipper) would be expected to have a larger body mass (and by how much), according to your model.

# Question 8 (15 points)

Now, let's focus just on our outcome, and how it is associated with species and island. Build a plot that shows the distribution of body mass (in grams) within each penguin species, and which also identifies the three islands (through the use of facets) in an attractive way. Your plot should have a meaningful title as well as useful axis and facet labels. Please restrict the plot to the 342 penguins with complete body mass information. 

Then, in a sentence or two, describe the results from your plot, and any reasonable conclusions you can draw from the plot about the relationships between island and species, and between body mass and the other two variables.

## On Grading 

Lab 04 will be graded on a 100 point scale, even though there are 105 available points, so that some people may score higher than 100. You will receive some credit for making a reasonable effort to build a good response to each question, and then additional credit if your response is (essentially) correct, your code is clean (in the sense that it is complete and can be run without modification to produce appropriate results) and you've written using complete and grammatically correct English sentences. 

A more detailed grading rubric will accompany the answer sketch.
