---
title: "431 Class 03 Demo: Exploring the Short Survey"
author: "Thomas E. Love, Ph.D."
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
    code_folding: show
---

```{r setup R and load packages, message=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 70)

## add additional libraries/packages here, as needed
## leaving the tidyverse as the last package loaded
library(magrittr); library(tidyverse)
```

# Load the Data

We will read in the .csv file of the data using `read_csv` to turn the data frame into a nicely organized *tibble*. 

- Since we've carefully stored the data file in the same directory as our R Project, we can read it in directly.

```{r load in a data set and make a tibble}
## if you want to load in a data set called namebeta.csv
## and then create a tibble from it called namealpha
## then uncomment the next line by removing the #

day1 <- read_csv("surveyday1_2020.csv") 
```

# Key Verbs in the Tidyverse for Data Wrangling

## Print your tibble by typing its name

```{r}
day1
```

## What are the column names?

```{r}
names(day1)
```


## Use **select** to pick columns / variables from your tibble

```{r}
day1 %>%
    select(favcolor, haircut)
```

## Use **filter** to pick rows / subjects from your tibble

```{r}
day1 %>%
    filter(year == 2020)
```

## Use **count** to count the number of observations meeting a criterion

```{r}
day1 %>%
    count(favcolor == "red")
```

Or to provide a cross-classification:

```{r}
day1 %>%
    count(favcolor == "blue", glasses)
```

## Use **arrange** to arrange the rows of a tibble

```{r}
day1 %>%
    count(smoke) %>%
    arrange(desc(n))
```

## Add new variables with **mutate**

```{r}
day1 %>%
    mutate(guess_error = ageguess - lovetrueage) %>%
    select(ageguess, lovetrueage, guess_error) %>%
    summary()
```

## Get grouped summaries with **group_by** and **summarize**

```{r}
day1 %>%
    group_by(year) %>%
    summarize(n = n(), 
              average_sleep = mean(lastsleep),
              min_sleep = min(lastsleep)) 
```

Whoops - looks like we have some missing `lastsleep` values. We could filter our data to only include the subjects who provided data on that variable...

```{r}
day1 %>%
    filter(complete.cases(lastsleep)) %>%
    group_by(year) %>%
    summarize(n = n(), 
              average_sleep = mean(lastsleep),
              min_sleep = min(lastsleep)) 
```

# Using `ggplot`: A Histogram of Haircut Prices

## Default Version

```{r}
ggplot(data = day1, aes(x = haircut)) +
    geom_histogram()
```

## Improvements

1. We'll filter the rows of the `day1` tibble to include only those subjects who gave us a haircut price.

```{r}
day1 %>% 
    filter(complete.cases(haircut)) %>%
    ggplot(data = ., aes(x = haircut)) +
    geom_histogram()
```

2. We'll specify that R should create bins of width $10 (rather than the default, which creates 30 bins) for the haircut prices to fall in.

```{r}
day1 %>% 
    filter(complete.cases(haircut)) %>%
    ggplot(data = ., aes(x = haircut)) +
    geom_histogram(binwidth = 10)
```

3. We'll set the fill to be a better color - a nice resource for this is to google **Colors in R**. I'll pick "slateblue". We'll also color the outlines of the bars "white".

```{r}
day1 %>% 
    filter(complete.cases(haircut)) %>%
    ggplot(data = ., aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
                   fill = "slateblue", col = "white")
```

4. We'll build a main title, subtitle and proper axis titles.

```{r}
day1 %>% 
    filter(complete.cases(haircut)) %>%
    ggplot(data = ., aes(x = haircut)) +
    geom_histogram(binwidth = 10, 
                   fill = "slateblue", col = "white") +
    labs(x = "Price of Most Recent Haircut in $",
         y = "Number of Students",
         title = "Histogram of Haircut Prices",
         subtitle = "431 students from 2014 - 2020")
```

## Separate histograms for each year with faceting?

```{r}
day1 %>%
    filter(!is.na(haircut)) %>%
    ggplot(., aes(x = haircut)) +
        geom_histogram(binwidth = 10, 
                       fill = "slateblue", color = "white") +
        guides(fill = FALSE) +
        labs(x = "Price of Last Haircut in $", 
        		 y = "Number of Students") +
        facet_wrap(~ year)
```

## Building a Comparison Boxplot

We could use a comparison boxplot. A trick here is to specify `year` as a factor...

```{r}
day1 %>%
    filter(complete.cases(haircut)) %>%
    ggplot(data = ., 
           aes(x = factor(year), y = haircut)) +
    geom_boxplot()
```


# Numerical Summaries

## Detailed Numerical Summary of Haircut Prices

```{r}
day1 %>%
    select(haircut) %>%
    summary()
```

which can also be done with

```{r}
summary(day1$haircut)
```

The `mosaic` package has a useful `favstats` function...

```{r}
mosaic::favstats(day1$haircut)
```

But to get this in a pipeline, you'd need the `%$%` operator from the `magrittr` package...

```{r}
day1 %$%
    mosaic::favstats(haircut)
```

The `psych` package has a useful `describe` function...

```{r}
day1 %$%
    psych::describe(haircut)
```

The `Hmisc` package also has a useful `describe` function...

```{r}
day1 %$%
    Hmisc::describe(haircut)
```

## Numerical Summary by Year?

```{r}
day1 %>% 
    filter(!is.na(haircut)) %>%
    group_by(year) %>%
    summarize(n = n(), mean = mean(haircut), 
              sd = sd(haircut), median = median(haircut))
```

# Using `ggplot` scatterplots

## What is the relationship between 431 students' pulse rate and hours of sleep the prior night?

Here, we're looking at two quantitative variables. A scatterplot is usually the best choice.

```{r}
ggplot(data = day1, aes(x = lastsleep, y = pulse)) +
    geom_point()
```

## Improving the Scatterplot

Let's filter to include only those cases with known `pulse` and known `lastsleep`, and also add a line from a linear regression model to predict pulse rate on the basis of hours of sleep the prior night.

```{r}
day1 %>%
    filter(complete.cases(pulse, lastsleep)) %>%
    ggplot(data = ., aes(x = lastsleep, y = pulse)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, col = "red")
```

Let's look at the 2019 data only, and fit a curved (loess) smooth to predict pulse rate on the basis of hours of sleep the prior night. We'll also add a title and subtitle and retitle the axes

```{r}
day1 %>%
    filter(year == "2019") %>%
    filter(complete.cases(pulse, lastsleep)) %>%
    ggplot(data = ., aes(x = lastsleep, y = pulse)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(title = "Pulse Rate as a Function of Hours of Sleep Last Night",
         subtitle = "with fitted loess smooth, students in the 2019 class",
         x = "Hours of Sleep Last Night",
         y = "Pulse Rate in Class Today")
```

We could instead restrict ourselves to a linear model for the 2019 group.

```{r}
day1 %>%
    filter(year == "2019") %>%
    filter(complete.cases(pulse, lastsleep)) %>%
    ggplot(data = ., aes(x = lastsleep, y = pulse)) +
    geom_point() +
    geom_smooth(method = "lm", col = "red") +
    labs(title = "Pulse Rate as a Function of Hours of Sleep Last Night",
         subtitle = "with fitted linear model, students in the 2019 class",
         x = "Hours of Sleep Last Night",
         y = "Pulse Rate in Class Today")
```

The correlation of `lastsleep` and `pulse` is likely to be of some interest. Note the use of both the `%>%` and `%$%` pipes in this case.

```{r}
day1 %>%
    filter(year == "2019") %>%
    filter(complete.cases(pulse, lastsleep)) %$%
    cor(pulse, lastsleep)
```

## Does the linear model change much by year?

Here's the plot, color coding the models by year (note the use of the group as well as the color aesthetic here), and also incorporating the black-and-white theme, rather than the default.

```{r}
day1 %>%
    filter(complete.cases(pulse, lastsleep)) %>%
    ggplot(., aes(x = lastsleep, y = pulse, 
    							color = factor(year),
    							group = factor(year))) +
    geom_point() + 
    geom_smooth(method = "lm", se = FALSE) +
	  theme_bw()
```

Here's the same plot of the models alone, and not showing the data (commenting out the line of code that draws the points.) We'll also improve the labeling.

```{r}
day1 %>%
    filter(complete.cases(pulse, lastsleep)) %>%
    ggplot(., aes(x = lastsleep, y = pulse, 
    							color = factor(year),
    							group = factor(year))) +
 #   geom_point() + 
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = "Pulse Rate and Hours of Sleep, 2014-2019",
         x = "Hours of Sleep Last Night",
         y = "Pulse Rate in Class Today") +
	  theme_bw()
```

## Faceting a Scatterplot

Here's the same basic information, but faceted by year.

```{r}
day1 %>%
    filter(complete.cases(pulse, lastsleep)) %>%
    ggplot(data = ., aes(x = lastsleep, y = pulse, 
    							group = factor(year))) +
    geom_point() + 
    geom_smooth(method = "lm", color = "red") +
    facet_wrap(~ year) +
    labs(title = "Pulse Rate and Hours of Sleep, 2014-2019",
         x = "Hours of Sleep Last Night",
         y = "Pulse Rate in Class Today") +
    theme_bw()
```