---
title: "Live Demo for Class 03"
author: "Thomas E. Love"
date: "`r Sys.Date()`"
output:
  html_document: 
    toc: yes
    code_folding: show
---

```{r setup R and load packages, message=FALSE}
knitr::opts_chunk$set(comment=NA)
options(width = 70)

## add additional libraries/packages here, as needed
## leaving the tidyverse as the last package loaded
library(tidyverse)
```

Loading in the data ...

```{r load in a data set and make a tibble}
## if you want to load in a data set called namebeta.csv
## and then create a tibble from it called namealpha
## then uncomment the next line by removing the #

day1 <- read_csv("surveyday1_2020.csv") 
```

# Looking at The Data

```{r}
day1
```
## Variable Names

```{r}
names(day1)
```

## Summary of the variables

```{r}
summary(day1)
```

## Table of glasses and english

```{r}
table(day1$glasses, day1$english)
```

## Table of english and sex

```{r}
table(day1$english, day1$sex)
```

```{r}
library(magrittr)

day1 %$% table(glasses, english)
```

```{r}
library(janitor)

day1 %>% tabyl(glasses, english)
```

# A Few Key Verbs

## filter

filter picks out rows in the data frame

```{r}
day1
```

```{r}
day1 %>% filter(glasses == "y")
```

```{r}
day1 %>% filter(glasses == "y", english == "n")
```

```{r}
day1 %>% filter(glasses == "y", favcolor == "blue") %>%
    count()
```

```{r}
day1 %>% filter(glasses == "y", favcolor == "blue") %>%
    count(year)
```

```{r}
day1 %>% filter(english == "y", favcolor == "blue") %>%
    count(year)
```

## select

select is used to pick out columns (variables) that we want to use

```{r}
day1 %>% select(smoke, favcolor, lastsleep)
```

Three key verbs so far: filter, select, count

More to come.

# Histogram of Last Night's Sleep

I build plots using the `ggplot2` package, which is part of the tidyverse. `ggplot2` has a function called `ggplot()`.

```{r}
ggplot(data = day1, aes(x = lastsleep)) +
    geom_histogram()
```

## Improve the Histogram

```{r}
ggplot(data = day1, aes(x = lastsleep)) +
    geom_histogram(binwidth = 1, fill = "royalblue", col = "yellow") +
    labs(title = "Histogram of Sleep for 431 Students")
```

## Boxplot of Sleep by English

```{r}
ggplot(data = day1, aes(x = english, y = lastsleep)) +
    geom_boxplot()
```

Get rid of the missing values...

```{r}
day1 %>% 
    filter(complete.cases(english, lastsleep)) %>%
    ggplot(data = ., aes(x = english, y = lastsleep)) +
    geom_boxplot()
```

## Faceted Histogram of Sleep by English

```{r}
ggplot(data = day1, aes(x = lastsleep)) +
    geom_histogram(binwidth = 1) +
    facet_wrap(~ english)
```

## Faceted Histogram of Sleep by Year

```{r}
ggplot(data = day1, aes(x = lastsleep)) +
    geom_histogram(binwidth = 1) +
    facet_wrap(~ year)
```

## Boxplot of Sleep by Year

```{r}
ggplot(data = day1, aes(x = factor(year), y = lastsleep)) +
    geom_boxplot()
```

## Scatterplot of Sleep by Height

```{r}
ggplot(data = day1, aes(x = height.in, y = lastsleep)) +
    geom_point()
```

## Add linear smooth and jitter points a little

```{r}
ggplot(data = day1, aes(x = height.in, y = lastsleep)) +
    geom_jitter(pch = 1) +
    geom_smooth(method = "lm") +
    labs(title = "Sleep and Height in 431 students",
         subtitle = "jittered points, with Linear fit")
```

## Add loess smooth

```{r}
ggplot(data = day1, aes(x = height.in, y = lastsleep)) +
    geom_point(pch = 1) +
    geom_smooth(method = "loess") +
    labs(x = "New title for X axis")
```

