---
title: "431 Lab 03"
date: "Last Edited `r Sys.time()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```


# General Instructions

Lab 03 is divided into 10 questions (including a short essay in Question 10.) 

Be sure to respond to each question by the deadline posted on the [Course Calendar](https://thomaselove.github.io/431/calendar.html). You are welcome (encouraged, really) to discuss Lab 03 with Dr. Love, the teaching assistants and even your colleagues, but your answer must be prepared by you alone. Don't be afraid to ask questions, either via [Piazza](https://piazza.com/case/fall2020/pqhs431) (use the lab03 folder), at [TA office hours](https://thomaselove.github.io/431/contact.html) or before/after class.

## An R Markdown Template for Lab 03

We have provided a useful R Markdown document template for this assignment called `YOURNAME-Lab03.Rmd` that you are welcome to use to help guide your work, but this is not mandatory.

- The `YOURNAME-Lab03.Rmd` file is available in the [main folder for Lab 03](https://github.com/THOMASELOVE/431-2020/tree/master/labs/lab03), and also on the [Data and Code repository](https://github.com/THOMASELOVE/431-data) for the course. 

The minimal setup for this assignment involves loading two relevant R packages, as shown below.

```{r}
library(MASS); library(tidyverse)
## make sure these libraries are installed in R
## always need tidyverse, also need MASS for Lab 03
```

## Data Description for Questions 1-7

The data come from the `faithful` data frame contained in the base installation of R, specifically, within the `datasets` package. These data describe the duration of an eruption for the Old Faithful geyser in Wyoming's Yellowstone National Park, as well as the waiting time until the next eruption. 

The commands below place the `faithful` data frame in a tibble called `lab03`, then display the tibble. We encourage you to use this `lab03` tibble to develop your answers to Questions 1-10.

```
lab03 <- tibble(faithful)
lab03
```

Use the command `?faithful` to see some additional details about these data. The `lab03` tibble contains observations on two variables:

- `eruptions` = eruption time in minutes
- `waiting` = waiting time to next eruption, also in minutes

# Question 1

Plot a histogram or other summary plot which meaningfully describes the distribution of the waiting times. Be sure it is very clearly labeled.

# Question 2 (on the center of the distribution)

What appears to be a typical waiting time?  Compare the mean, median and 80\% trimmed mean (mean of the middle 80% of the observed waiting times.)

# Question 3 (on the spread of the distribution)

What is the inter-quartile range, and how does it compare to the standard deviation?

# Question 4 (on the shape of the distribution)

Describe the shape of the distribution in a few complete sentences. You should address the following questions in your response.

- Is the distribution multi-modal or unimodal? How do you know?
- Is the distribution skewed (and if so, in which direction) or is it essentially symmetric? How do you know?
- Are there any unusual (outlier) values in the distribution, and if so, what are they? 

# Question 5

In light of your work in Questions 1-4, would a model using the Normal distribution be an appropriate way to summarize the waiting time data? Why or why not?

# Question 6

Plot a scatterplot of the waiting times (y-axis) vs. the eruption durations (x-axis) and be sure your plot is very clearly labeled. Include the result of fitting a straight line regression model in your plot. Then describe your general impression of the plot: what sort of relationship do you see?

# Question 7

Does a linear model seem like an appropriate thing to use in attempting to predict the waiting time given the most recent eruption duration, based on these data? Why or why not? Your response should reference the Pearson correlation coefficient for the plot you've developed, but should also react to the appearance of the plot beyond what you learn from the correlation coefficient.

# Question 8

Investigate question 6 again, but now use the `geyser` data in the `MASS` package, and compare your results to what you obtained originally, in an appropriate way. 

# Question 9

Investigate question 7 again, but now use the `geyser` data.

## Important Hints for Questions 8-9

1. Be sure to use the command `?MASS::geyser` to see some additional details about these data, and to learn more about how they differ from the data in our original data set. 
2. There is at least one key difference between the variables available in `geyser` and those which you dealt with in the `faithful` data. For full credit, you must specify that difference (and describe its impact) in your response.
3. The commands below will place the relevant data for Questions 8 and 9 in the `lab03extra` tibble.

```
lab03extra <- tibble(MASS::geyser)
lab03extra
```

# Question 10

In your reading of Jeff Leek's book so far (you should have completed at least Sections 1-5 and 9-12), what is the most important thing you've learned? 

- This could be an idea that has stuck with you, a phrase or sentence that clarified something of importance, or a tip for working with data that you want to remember. 

In a short essay (or perhaps 100-150 words), identify the relevant passage from Leek's book appropriately (be sure to specify the section of the book and perhaps provide a short quotation), and then provide a brief argument as to why this particular thing is something you value, and how it might apply to your work.

# Submitting your Response

Build your entire response as an R Markdown file (if you like you are free to use the `YOURNAME-lab03.Rmd` template we have provided but this is not mandatory.) Then use the Knit button in RStudio to create the resulting HTML document. Submit both your revised R Markdown file and the HTML output file to [Canvas in the Lab 03 section of the Assignments folder](https://canvas.case.edu) by the deadline specified in [the Course Calendar](https://thomaselove.github.io/431/calendar.html).

# On Grading 

Lab 03 will be graded on a 0-100 scale, with question 10 weighted more heavily than the other questions. You will receive some credit for making a reasonable effort to build a good response to each question, and then additional credit if your response is (essentially) correct, your code is clean and you've written any text using complete English sentences. 

A more detailed grading rubric will accompany the answer sketch.

