---
title: "431 Class 17"
author: "thomaselove.github.io/431"
date: "2020-10-22"
output:
  beamer_presentation:
    theme: "Madrid"
    fonttheme: "structurebold"
    colortheme: "whale"
    fig_height: 5.5
    fig_caption: false
---

```{r set-options, echo=FALSE, cache=FALSE, message = FALSE}
options(width = 55)
```

## Today's Agenda

Comparing Population Means with the Analysis of Variance

- Indicator Variable Regression Analysis
  - Interpreting the ANOVA table
  - ANOVA assumptions and the Kruskal-Wallis test
- The Problem of Multiple Comparisons
  - Bonferroni pairwise testing
  - Tukey HSD pairwise comparisons

## Today's Setup and Data

```{r load_packages, message = FALSE}
knitr::opts_chunk$set(comment = NA) 
options(dplyr.summarise.inform = FALSE) 

library(readxl) # to read in an .xlsx file
library(ggrepel) # to help label residual plots
library(patchwork)
library(knitr)
library(magrittr)
library(janitor)
library(broom)
library(tidyverse)

theme_set(theme_bw())
```

## Today's Data (`ohio_2020.xlsx`)

`ohio_2020.xlsx` rows describe one of Ohio's 88 counties in terms of:

- `FIPS` code (basically an identifier for mapping) 
- `state` and `county` name
- health outcomes (standardized: more positive means **better** outcomes, because we've taken the negative of the Z score CHR provides)
- health behavior ranking (1-88, we'll divide into 4 groups)
- clinical care ranking (1-88, we'll split into 3 groups)
- proportion of county residents who live in rural areas
- median income, in dollars
- proportion of votes in the 2016 Presidential Election for Pres. Trump

### Sources (these bullets are links)

- [County Health Rankings](https://www.countyhealthrankings.org/app/ohio/2020/downloads) (2020 Ohio Data)
- [Wikipedia for 2016 Election Results](https://en.wikipedia.org/wiki/2016_United_States_presidential_election_in_Ohio#By_county)

## Importing the Data / Creating some Factors

```{r}
ohio20 <- read_xlsx("data/ohio_2020.xlsx") %>%
  mutate(behavior = Hmisc::cut2(rk_behavior, g = 4),
         clin_care = Hmisc::cut2(rk_clin_care, g = 3)) %>%
  mutate(behavior = fct_recode(behavior,
            "Best" = "[ 1,23)", "High" = "[23,45)",
            "Low" = "[45,67)", "Worst" = "[67,88]")) %>%
  mutate(clin_care = fct_recode(clin_care,
            "Strong" = "[ 1,31)", "Middle" = "[31,60)",
            "Weak" = "[60,88]")) %>%
  select(FIPS, state, county, outcomes, behavior, clin_care, 
         everything())
```

## A Quick Look at the Data

```{r}
ohio20 %>% filter(county == "Cuyahoga") %>%
  select(FIPS, county, outcomes, behavior, clin_care) 
```

```{r, fig.height = 2}
ggplot(ohio20, aes(x = "", y = outcomes)) + geom_violin() +
  geom_boxplot(width = 0.4) + coord_flip() + labs(x = "")
```

## Use `inspect` to inspect the data frame?

```{r, message = FALSE}
ohio20 %>% select(outcomes, behavior, clin_care) %>%
  mosaic::inspect()
```

## Key Measure Details

- **outcomes** = quantity that describes the county's premature death and quality of life results, weighted equally and standardized (z scores).
  - Higher (more positive) values indicate better outcomes in this county.
- **behavior** = (Best/High/Low/Worst) reflecting adult smoking, obesity, food environment, inactivity, exercise, drinking, alcohol-related driving deaths, sexually tranmitted infections and teen births. 
  - Counties in the Best group had the best behavior results.
- **clin_care** = (Strong/Middle/Weak) reflects rates of uninsured, care providers, preventable hospital stays, diabetes monitoring and mammography screening.
  - Strong means that clinical care is strong in this county.

## Today's Planned Analytic Questions

1. How do average health outcomes vary across groups of counties defined by health behavior?

2. Do groups of counties defined by clinical care show substantial differences in average health outcomes?

# Health Outcomes compared across Health Behavior Groups

## Question 1

Do average health outcomes differ by health behavior?

```{r, echo = FALSE}
ggplot(ohio20, aes(x = behavior, y = outcomes, 
                   fill = behavior)) +
  geom_violin(alpha = 0.25) +
  geom_boxplot(width = 0.25) +
  guides(fill = FALSE) + 
  scale_fill_brewer(palette = "Spectral", direction = -1) +
  theme_bw() +
  labs(x = "Health Behavior Group", 
       y = "Health Outcomes (higher = better health)",
       title = "Health Outcomes across Behavior Groups",
       subtitle = "Ohio's 88 counties, 2020 County Health Rankings",
       caption = "Source: https://www.countyhealthrankings.org/app/ohio/2020/downloads")
```

## Question 1 Numerical Summaries

How do average health outcomes vary across groups of counties defined by health behavior?

```{r}
mosaic::favstats(outcomes ~ behavior, data = ohio20) %>%
  rename(na = missing) %>% knitr::kable(digits = 2)
```

Note that there is no missing data here.

## Analysis of Variance (ANOVA) testing: Question 1

Does the mean `outcomes` result differ detectably across the `behavior` groups?

$$
H_0: \mu_{Best} = \mu_{High} = \mu_{Low} = \mu_{Worst} \mbox{ vs. } H_A: \mbox{At least one } \mu \mbox{ is different.}
$$

To test this set of hypotheses, we will build a linear model to predict each county's outcome based on what behavior group the county is in.

- We then look at whether the `behavior` group effect has a statistically detectable impact on the model's predictions of `outcomes`. 

## Building the Linear Model: Question 1

Can we detect differences in the population means of `outcomes` across the four `behavior` groups, using a 10% significance level? 

```{r}
model_one <- lm(outcomes ~ behavior, data = ohio20)
tidy(model_one, conf.int = 0.90) %>% 
  select(term, estimate, std.error, 
         conf.low, conf.high, p.value) %>% kable(dig = 2)
```

How do we interpret this result?

## Interpreting the Indicator Variables

The regression model (`model_one`) equation is 

```
outcomes = 0.96 - 0.71 behaviorHigh 
                    - 1.14 behaviorLow 
                      - 2.01 behaviorWorst
```

What do the indicator variables mean?

group  | `behaviorHigh` | `behaviorLow` | `behaviorWorst`
----: | :-------: | :--------: | :--------:
Best  | 0 | 0 | 0
High  | 1 | 0 | 0
Low   | 0 | 1 | 0
Worst | 0 | 0 | 1

- So what is the predicted `outcomes` score for a county in the High behavior group, according to this model?

## Interpreting the Indicator Variables

The regression model (`model_one`) equation is 

```
outcomes = 0.96 - 0.71 behaviorHigh 
                    - 1.14 behaviorLow 
                      - 2.01 behaviorWorst
```

What predictions does the model make?

group  | `High` | `Low` | `Worst` | Prediction
----: | :-----: | :------: | :------: | --------------
Best  | 0 | 0 | 0 | 0.96
High  | 1 | 0 | 0 | 0.96 - 0.71 = 0.25
Low   | 0 | 1 | 0 | 0.96 - 1.14 = -0.18
Worst | 0 | 0 | 1 | 0.96 - 2.01 = -1.05

Do these predictions make sense?

## Interpreting the Indicator Variables

The regression model (`model_one`) equation is 

```
outcomes = 0.96 - 0.71 behaviorHigh 
                    - 1.14 behaviorLow 
                      - 2.01 behaviorWorst
```

Sample means are...

```{r}
ohio20 %>% group_by(behavior) %>%
  summarize(n = n(), mean = round_half_up(mean(outcomes),2)) %>% kable()
```


## ANOVA for the Linear Model: Question 1

Are there statistically detectable differences in mean outcome across the behavior group means?

$$
H_0: \mu_{Best} = \mu_{High} = \mu_{Low} = \mu_{Worst} \mbox{ vs. } H_A: \mbox{At least one } \mu \mbox{ is different.}
$$

```{r}
anova(model_one)
```

## So, what's in the ANOVA table? (df)

The ANOVA table reports here on a single **factor** (behavior group) with 4 levels, and on the residual variation in health **outcomes**.

```{r, echo = FALSE}
anova(model_one)[1:4]
```

**Degrees of Freedom** (df) is an index of sample size...

- df for our factor (behavior) is one less than the number of categories. We have four behavior groups, so 3 degrees of freedom.
- Adding df(behavior) + df(Residuals) = 3 + 84 = 87 = df(Total), one less than the number of observations (counties) in Ohio.
- *n* observations and *g* groups yield $n - g$ residual df in a one-factor ANOVA table.

## So, what's in the ANOVA table? (Sum of Squares)

```{r, echo = FALSE}
anova(model_one)[1:4]
```

**Sum of Squares** (`Sum Sq`, or SS) is an index of variation...

- SS(factor), here SS(`behavior`) measures the amount of variation accounted for by the `behavior` groups in our `model_one`.
- The total variation in `outcomes` to be explained by the model is SS(factor) + SS(Residuals) = SS(Total) in a one-factor ANOVA table.
- We describe the proportion of variation explained by a one-factor ANOVA model with $\eta^2$ ("eta-squared": same as Multiple $R^2$)

$$
\eta^2 = \frac{SS(\mbox{behavior})}{SS(\mbox{Total})} = \frac{46.421}{46.421+22.519} = \frac{46.421}{68.94} \approx 0.673
$$

## So, what's in the ANOVA table? (MS and F)

```{r, echo = FALSE}
anova(model_one)[1:4]
```

**Mean Square** (`Mean Sq`, or MS) = Sum of Squares / df

$$
MS(\mbox{behavior}) = \frac{SS(\mbox{behavior})}{df(\mbox{behavior})} = \frac{46.421}{3} \approx 15.4736
$$

- MS(Residuals) estimates the **residual variance**, the square of the residual standard deviation (residual standard error in earlier work).
- The ratio of MS values is the ANOVA **F value**.

$$
{\mbox{ANOVA }} F = \frac{MS(\mbox{behavior})}{MS(\mbox{Residuals})} = \frac{15.4736}{0.2681} \approx 57.718
$$


## So, what's in the ANOVA table? (p value)

```{r}
tidy(anova(model_one)) %>% kable(dig = 3)
```

- The *p* value is derived from the ANOVA F statistic, as compared to the F distribution.
- Which F distribution is specified by the two degrees of freedom values, as the F table is indexed by both a numerator and a denominator df.

```{r}
pf(57.718, df1 = 3, df2 = 84, lower.tail = FALSE)
```

## Alternative ways to show ANOVA results

```{r}
glance(model_one) %>% select(r.squared, statistic, df, df.residual, p.value)
```

```{r}
summary(aov(model_one))
```

So, what's the conclusion? Is this a surprise?

## ANOVA Assumptions

The assumptions behind analysis of variance are those of a linear model. Of specific interest are:

- The samples obtained from each group are independent.
- Ideally, the samples from each group are a random sample from the population described by that group.
- In the population, the variance of the outcome in each group is equal. (This is less of an issue if our study involves a balanced design.)
- In the population, we have Normal distributions of the outcome in each group.

Happily, the ANOVA F test is fairly robust to violations of the Normality assumption.

## Residual Plots for `model_one`

```{r, echo = FALSE}
aug_one <- augment(model_one, ohio20)

p1 <- ggplot(aug_one, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = F,
              lty = "dashed", col = "red") +
  geom_text_repel(data = aug_one %>% 
                    slice_max(abs(.resid), n = 3), 
                  aes(label = county)) +
  labs(title = "model_one Residuals vs. Fitted",
       x = "Fitted Value from model_one",
       y = "Residuals from model_one")

p2 <- ggplot(aug_one, aes(sample = .resid)) +
  geom_qq() + geom_qq_line(col = "red") + 
  labs(title = "model_one Residuals",
       y = "")

p3 <- ggplot(aug_one, aes(y = .resid, x = "")) +
  geom_violin(fill = "aquamarine") +
  geom_boxplot(width = 0.5) + 
  labs(y = "", x = "")

p1 + p2 + p3 + plot_layout(widths = c(5, 4, 1))
```

## Can we avoid assuming equal population variances?

Yes, but this isn't exciting if we have a balanced design.

```{r}
oneway.test(outcomes ~ behavior, data = ohio20)
```

- Note that this approach uses a fractional degrees of freedom calculation in the denominator.

## The Kruskal-Wallis Test

If you thought the data were severely skewed, you might try:

```{r}
kruskal.test(outcomes ~ behavior, data = ohio20)
```

- $H_0$: The four `behavior` groups have the same center to their `outcomes` distributions.
- $H_A$: At least one group has a shifted distribution, with a different center to its `outcomes`.

What would be the conclusion here?

# Health Outcomes compared across Clinical Care Groups

## Question 2

Do groups of counties defined by clinical care show meaningful differences in average health outcomes?

```{r, echo = FALSE}
ggplot(ohio20, aes(x = clin_care, y = outcomes, 
                   fill = clin_care)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.25, notch = TRUE, 
               col = c("white", "black", "black")) +
  guides(fill = FALSE) + 
  scale_fill_viridis_d(option = "C") +
  theme_bw() +
  labs(x = "Clinical Care Ranking (groups)", 
       y = "Health Outcomes (higher = better health)",
       title = "Health Outcomes across County Clinical Care Ranking",
       subtitle = "Ohio's 88 counties, 2020 County Health Rankings",
       caption = "Source: https://www.countyhealthrankings.org/app/ohio/2020/downloads")
```

## Question 2 Numerical Summaries

Do groups of counties defined by clinical care show meaningful differences in average health outcomes?

```{r}
mosaic::favstats(outcomes ~ clin_care, data = ohio20) %>%
  rename(na = missing) %>% knitr::kable(digits = 2)
```

## Question 2 Analysis of Variance

```{r}
model_two <- lm(outcomes ~ clin_care, data = ohio20)

anova(model_two)
```

## Residual Plots for `model_two`

```{r, echo = FALSE}
aug_two <- augment(model_two, ohio20)

p1 <- ggplot(aug_two, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = F,
              lty = "dashed", col = "red") +
  geom_text_repel(data = aug_two %>% 
                    slice_max(abs(.resid), n = 3), 
                  aes(label = county)) +
  labs(title = "model_two Residuals vs. Fitted",
       x = "Fitted Value from model_two",
       y = "Residuals from model_two")

p2 <- ggplot(aug_two, aes(sample = .resid)) +
  geom_qq() + geom_qq_line(col = "red") + 
  labs(title = "model_two Residuals",
       y = "")

p3 <- ggplot(aug_two, aes(y = .resid, x = "")) +
  geom_violin(fill = "aquamarine") +
  geom_boxplot(width = 0.5) + 
  labs(y = "", x = "")

p1 + p2 + p3 + plot_layout(widths = c(5, 4, 1))
```

## Question 2 Kruskal-Wallis test

```{r}
kruskal.test(outcomes ~ clin_care, data = ohio20)
```


## K-Sample Study Design, Comparing Means

1. What is the outcome under study?
2. What are the (in this case, $K \geq 2$) treatment/exposure groups?
3. Were the data in fact collected using independent samples?
4. Are the data random samples from the population(s) of interest? Or is there at least
a reasonable argument for generalizing from the samples to the population(s)?
5. What is the significance level (or, the confidence level) we require?
6. Are we doing one-sided or two-sided testing? (usually 2-sided)
7. What does the distribution of each individual sample tell us about which inferential procedure to use?
8. Are there statistically meaningful differences between population means?
9. If an overall test rejects the null, can we identify pairwise comparisons of means that show detectable differences using an appropriate procedure that protects against Type I error expansion due to multiple comparisons?

# Multiple Comparisons

## What's Left to do? (Multiple Comparisons)

9. If an overall test rejects the null, can we identify pairwise comparisons of means that show detectable differences using an appropriate procedure that protects against Type I error expansion due to multiple comparisons?

Yes. There are two methods we'll study to identify specific pairs of means where we have statistically detectable differences, while dealing with the problem of multiple comparisons.

- Bonferroni pairwise comparisons
- Tukey's HSD (Honestly Significant Differences) approach

## Compare `behavior` group means of `outcomes`?

ANOVA tells is that there is strong evidence that they aren't all the same. Which ones are different from which?

```{r}
anova(lm(outcomes ~ behavior, data = ohio20))
```

Is, for example, Best detectably different from Worst?

## Could we just run a bunch of t tests?

This approach assumes that you need to make no adjustment for the fact that you are doing multiple comparisons, simultaneously.

```{r}
pairwise.t.test(ohio20$outcomes, ohio20$behavior, 
                p.adjust.method = "none")
```

## The problem of Multiple Comparisons

- The more comparisons you do simultaneously, the more likely you are to make an error.

In the worst case scenario, suppose you do two tests - first A vs. B and then A vs. C, each at the $\alpha = 0.10$ level.

- What is the combined error rate across those two t tests?

## The problem of Multiple Comparisons

In the worst case scenario, suppose you do two tests - first A vs. B and then A vs. C, each at the $\alpha = 0.10$ level.

- What is the combined error rate across those two t tests?

Run the first test. Make a Type I error 10% of the time.

A vs B Type I error | Probability
-----------: | -----------
Yes | 0.1
No  | 0.9

Now, run the second test. Assume (perhaps wrongly) that comparing A to C is independent of your A-B test result. What is the error rate now?


## The problem of Multiple Comparisons

In the worst case scenario, suppose you do two tests - first A vs. B and then A vs. C, each at the $\alpha = 0.10$ level.

- What is the combined error rate across those two t tests?

Assuming there is a 10% chance of making an error in either test, independently ...

-- | Error in A vs. C  | No Error | Total
----------------------: | --------: | --------: | ----:
Type I error in A vs. B | 0.01 | 0.09 | 0.10
No Type I error in A-B  | 0.09 | 0.81 | 0.90
Total                   | 0.10 | 0.90 | 1.00

So you will make an error in the A-B or A-C comparison **19%** of the time, rather than the nominal $\alpha = 0.10$ error rate.

## But in our case, we're building SIX tests

1. Best vs. High
2. Best vs. Low
3. Best vs. Worst
4. High vs. Low
5. High vs. Worst
6. Low vs. Worst

and if they were independent, and each done at a 5% error rate, we could still wind up with an error rate of 

$.05 + (.95)(.05) + (.95)(.95)(.05) + (.95)^3(.05) + (.95)^4(.05) + (.95)^5(.05)$ = .265

Or worse, if they're not independent.

## The Bonferroni Method

If we do 6 tests, we could reduce the necessary $\alpha$ to 0.05 / 6 = 0.0083 and that maintains an error rate no higher than $\alpha = 0.05$ across the 6 tests.

- Or, R can adjust the *p* values directly...

```{r}
pairwise.t.test(ohio20$outcomes, ohio20$behavior, 
                p.adjust.method = "bonferroni")
```

We still detect a meaningful difference between each pair of groups.

## Tukey Honestly Significant Differences (HSD)

Tukey's HSD approach is a better choice for pre-planned comparisons with a balanced (or nearly balanced) design. It provides confidence intervals and an adjusted *p* value for each comparison.

- Let's run some confidence intervals to yield an overall 99% confidence level, even with 6 tests...

```{r, eval = FALSE}
TukeyHSD(aov(lm(outcomes ~ behavior, data = ohio20)), 
         conf.level = 0.99, ordered = TRUE)
```

Output on the next slide...

## Tukey HSD Output

```{r, echo = FALSE}
TukeyHSD(aov(lm(outcomes ~ behavior, data = ohio20)), 
         conf.level = 0.99, ordered = TRUE)
```


## Tidying the Tukey HSD confidence intervals

```{r}
model_one <- lm(outcomes ~ behavior, data = ohio20)
tukey_one <- tidy(TukeyHSD(aov(model_one), 
                           ordered = TRUE, 
                           conf.level = 0.99))
tukey_one %>% rename(null = null.value) %>% kable(dig = 3)
```

## Plotting Your Tukey HSD intervals, Approach 1

```{r, echo = FALSE}
ggplot(tukey_one, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_text(aes(label = round(estimate,2)), nudge_x = -0.2) +
  theme_bw() +
  labs(x = "Contrast between Behavior Groups", 
       y = "Estimated Effect, with 99% Tukey HSD interval",
       title = "Estimated Effects, with Tukey HSD 99% Confidence Intervals",
       subtitle = "Comparing Outcomes by Behavior Group, ohio20 data")
```

## Code for Plot on Previous Slide

```{r, eval = FALSE}
ggplot(tukey_one, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_text(aes(label = round(estimate,2)), nudge_x = -0.2) +
  theme_bw() +
  labs(x = "Contrast between Behavior Groups", 
       y = "Estimated Effect, with 99% Tukey HSD interval",
       title = "Estimated Effects, with Tukey HSD 99% Confidence Intervals",
       subtitle = "Comparing Outcomes by Behavior Group, ohio20 data")
```

## Question 2: 90% Tukey HSD intervals, tidying

```{r}
model_two <- lm(outcomes ~ clin_care, data = ohio20)
tukey_two <- tidy(TukeyHSD(aov(model_two), 
                           ordered = TRUE, 
                           conf.level = 0.90))
tukey_two %>% select(-term, -null.value) %>% kable(dig = 3)
```

## Plotting Question 2 Tukey HSD intervals

```{r, echo = FALSE}
ggplot(tukey_two, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_crossbar(aes(ymin = conf.low, ymax = conf.high), 
                fatten = 1) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_text(aes(label = round(estimate,2)), nudge_y = 0.1) +
  theme_bw() +
  labs(x = "Contrast between Clinical Care Groups", 
       y = "Estimated Effect, with 90% Tukey HSD interval",
       title = "Estimated Effects, with Tukey HSD 90% Confidence Intervals",
       subtitle = "Comparing Outcomes by Clinical Care Group, ohio20 data")
```

## Code for Question 2 Tukey HSD plot

```{r, eval = FALSE}
ggplot(tukey_two, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_crossbar(aes(ymin = conf.low, ymax = conf.high), 
                fatten = 1) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_text(aes(label = round(estimate,2)), nudge_y = 0.1) +
  theme_bw() +
  labs(x = "Contrast between Clinical Care Groups", 
       y = "Estimated Effect, with 90% Tukey HSD interval",
       title = "Estimated Effects, with Tukey HSD 90% Confidence Intervals",
       subtitle = "Comparing Outcomes by Clinical Care Group, ohio20 data")
```

# Coming Up Next Time: Comparisons using Paired/Matched Samples

# FOR SELF-STUDY: ANOVA Examples about President Trump's 2016 Votes by County

## Question 3 (Education)

We have some additional variables in `ohio20`, specifically:

- `trump16` = proportion of the vote cast in 2016 in the county that went to President Trump
- `somecollege` = percentage of adults ages 25-44 with some post-secondary education in the county

Let's break Ohio's counties into 5 groups based on `somecollege`...

```{r}
ohio20 <- ohio20 %>% 
  mutate(trump16 = 100*trump16) %>%
  mutate(educ = Hmisc::cut2(somecollege, g = 5)) %>%
  mutate(educ = fct_recode(educ, "Least" = "[20.4,50.3)", 
          "Low" = "[50.3,54.3)", "Middle" = "[54.3,59.7)", 
          "High" = "[59.7,67.1)", "Most" = "[67.1,85.1]"))
```

Did President Trump's vote percentage in 2016 vary meaningfully across groups of counties defined by educational attainment?

## Trump 2016 % by Educational Attainment

```{r, echo = FALSE}
ggplot(ohio20, aes(x = educ, y = trump16, fill = educ)) +
  geom_violin(alpha = 0.25) +
  geom_boxplot(width = 0.25) +
  guides(fill = FALSE) + 
  scale_fill_brewer(palette = "Spectral", direction = -1) +
  theme_bw() +
  labs(x = "Education Group (2020 County Health Rankings)", 
       y = "Proportion of Vote for Trump in 2016 Election",
       title = "Proportion of Trump Vote by 'Some College' Group",
       subtitle = "Ohio's 88 counties")
```

## Numerical Comparison

```{r}
mosaic::favstats(trump16 ~ educ, data = ohio20) %>%
  rename(na = missing) %>% kable(digits = 2)
```

## Analysis of Variance (ANOVA) testing: Question 3

Does the mean `trump16` result differ detectably across the `educ` groups?

```{r}
model_3 <- lm(trump16 ~ educ, data = ohio20)

tidy(model_3, conf.int = 0.90) %>% 
  select(term, estimate, std.error, 
         conf.low, conf.high, p.value) %>% kable(dig = 2)
```

## ANOVA for the Linear Model: Question 3

```{r}
anova(model_3)
```

```{r}
glance(model_3) %>% 
  select(r.squared, statistic, df, df.residual, p.value)
```

So, what's the conclusion?

## Residual Plots for `model_3`

```{r, echo = FALSE}
aug_3 <- augment(model_3, ohio20)

p1 <- ggplot(aug_3, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = F,
              lty = "dashed", col = "red") +
  geom_text_repel(data = aug_3 %>% 
                    slice_max(abs(.resid), n = 3), 
                  aes(label = county)) +
  labs(title = "model_3 Residuals vs. Fitted",
       x = "Fitted Value from model_3",
       y = "Residuals from model_3")

p2 <- ggplot(aug_3, aes(sample = .resid)) +
  geom_qq() + geom_qq_line(col = "red") + 
  labs(title = "model_3 Residuals",
       y = "")

p3 <- ggplot(aug_3, aes(y = .resid, x = "")) +
  geom_violin(fill = "aquamarine") +
  geom_boxplot(width = 0.5) + 
  labs(y = "", x = "")

p1 + p2 + p3 + plot_layout(widths = c(5, 4, 1))
```

## Does Kruskal-Wallis give a very different result?

```{r}
kruskal.test(trump16 ~ educ, data = ohio20)
```

## Tukey HSD 90% confidence intervals: Example 3

```{r}
tukey_3 <- tidy(TukeyHSD(aov(model_3), 
                         ordered = TRUE, 
                         conf.level = 0.90))
tukey_3 %>% select(-null.value) %>% kable(dig = 3)
```

## Plotting Tukey HSD intervals for Example 3

```{r, echo = FALSE}
ggplot(tukey_3, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_label(aes(label = round_half_up(estimate,1))) +
  coord_flip() +
  labs(x = "Contrast between Education Groups", 
       y = "Estimated Effect, with 90% Tukey HSD interval",
       title = "Estimated Effects, with Tukey HSD 90% Confidence Intervals",
       subtitle = "Comparing Trump16 Vote % by Education Group, ohio20 data")
```

## Code for Previous Slide

```{r, eval = FALSE}
ggplot(tukey_3, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_label(aes(label = round_half_up(estimate,1))) +
  coord_flip() +
  labs(x = "Contrast between Education Groups", 
   y = "Estimated Effect, with 90% Tukey HSD interval",
   title = "Estimated Effects, with Tukey HSD 90% Confidence Intervals",
   subtitle = "Comparing Trump16 Vote % by Education Group, ohio20 data")
```

## Question 4 ()

Let's break Ohio's counties into 4 groups based on their median `income`...

```{r}
ohio20 <- ohio20 %>% 
  mutate(income = Hmisc::cut2(income, g = 4)) %>%
  mutate(income = fct_recode(income, "Lowest" = "[40416, 48792)", 
          "Low" = "[48792, 53904)", "High" = "[53904, 60828)", 
          "Highest" = "[60828,103536]"))
```

Did President Trump's vote percentage in 2016 vary meaningfully across income?

## Trump 2016 % by Income

```{r, echo = FALSE}
ggplot(ohio20, aes(x = income, y = trump16, fill = income)) +
  geom_violin(alpha = 0.25) +
  geom_boxplot(width = 0.25) +
  guides(fill = FALSE) + 
  scale_fill_brewer(palette = "Spectral", direction = -1) +
  theme_bw() +
  labs(x = "Income Group (2020 County Health Rankings)", 
       y = "Proportion of Vote for Trump in 2016 Election",
       title = "Proportion of Trump Vote by Income Group",
       subtitle = "Ohio's 88 counties")
```

## Numerical Comparison

```{r}
mosaic::favstats(trump16 ~ income, data = ohio20) %>%
  rename(na = missing) %>% kable(digits = 2)
```

## Analysis of Variance (ANOVA) testing

Does the mean `trump16` result differ detectably across the `income` groups?

```{r}
model_4 <- lm(trump16 ~ income, data = ohio20)

tidy(model_4, conf.int = 0.90) %>% 
  select(term, estimate, std.error, 
         conf.low, conf.high, p.value) %>% kable(dig = 2)
```

## ANOVA for the Linear Model

```{r}
anova(model_4)
```

```{r}
glance(model_4) %>% 
  select(r.squared, statistic, df, df.residual, p.value)
```

So, what's the conclusion?

## Residual Plots for `model_4`

```{r, echo = FALSE}
aug_4 <- augment(model_4, ohio20)

p1 <- ggplot(aug_4, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = F,
              lty = "dashed", col = "red") +
  geom_text_repel(data = aug_4 %>% 
                    slice_max(abs(.resid), n = 3), 
                  aes(label = county)) +
  labs(title = "model_4 Residuals vs. Fitted",
       x = "Fitted Value from model_4",
       y = "Residuals from model_4")

p2 <- ggplot(aug_4, aes(sample = .resid)) +
  geom_qq() + geom_qq_line(col = "red") + 
  labs(title = "model_4 Residuals",
       y = "")

p3 <- ggplot(aug_4, aes(y = .resid, x = "")) +
  geom_violin(fill = "aquamarine") +
  geom_boxplot(width = 0.5) + 
  labs(y = "", x = "")

p1 + p2 + p3 + plot_layout(widths = c(5, 4, 1))
```

## Does Kruskal-Wallis give a different result?

```{r}
kruskal.test(trump16 ~ income, data = ohio20)
```

## Tukey HSD 90% confidence intervals: Income Groups

```{r}
tukey_4 <- tidy(TukeyHSD(aov(model_4), 
                         ordered = TRUE, 
                         conf.level = 0.90))
tukey_4 %>% select(-null.value) %>% kable(dig = 3)
```

## Plotting Tukey HSD intervals (Income Groups)

```{r, echo = FALSE}
ggplot(tukey_4, aes(x = reorder(contrast, -estimate), 
                      y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) + 
  geom_hline(yintercept = 0, col = "red", 
             linetype = "dashed") +
  geom_label(aes(label = round_half_up(estimate,2))) +
  coord_flip() +
  labs(x = "Contrast between Income Groups", 
       y = "Estimated Effect, with 90% Tukey HSD interval",
       title = "Estimated Effects, with Tukey HSD 90% Confidence Intervals",
       subtitle = "Comparing Trump16 Vote % by Income Group, ohio20 data")
```

## K-Sample Study Design, Comparing Means

1. What is the outcome under study?
2. What are the (in this case, $K \geq 2$) treatment/exposure groups?
3. Were the data in fact collected using independent samples?
4. Are the data random samples from the population(s) of interest? Or is there at least
a reasonable argument for generalizing from the samples to the population(s)?
5. What is the significance level (or, the confidence level) we require?
6. Are we doing one-sided or two-sided testing? (usually 2-sided)
7. What does the distribution of each individual sample tell us about which inferential procedure to use?
8. Are there statistically meaningful differences between population means?
9. If an overall test rejects the null, can we identify pairwise comparisons of means that show detectable differences using an appropriate procedure that protects against Type I error expansion due to multiple comparisons?

